\chapter{Conclusions}
In system optimization 
it is not possible to apply a general optimization algorithm that works efficiently 
on all problems.
What algorithm should be used depends on the properties of the cost function,
such as the number of independent parameters,
the continuity of the cost function and its derivatives,
and the existence of local minima.
Thus a variety of optimization algorithms is needed.
To address this need, GenOpt has a library with different optimization algorithms
and an optimization algorithm interface that users can use to implement 
their own optimization algorithm if desired.\\

The fact that analytical properties of the cost function are unavailable 
for the class of optimization problems that GenOpt has been developed for
makes it possible to separate optimization and function evaluation.
Therefore, 
GenOpt has a simulation program interface that allows coupling any 
program that exchanges input and output using text files.
Hence,
users are not restricted to using a special program for evaluating the cost function. 
Rather, they can use the simulation program they are already using for their 
system design and development. 
Thus, the system can be optimized with little additional effort.\\

This open environment not only allows coupling any simulation program and 
implementing special purpose algorithms, 
but it also allows sharing algorithms among users.
This makes it possible to extend the algorithm library 
and thus extend GenOpt's applicability.\\

% =========================================================
\chapter{Acknowledgments}
The development of GenOpt was sponsored by grants from 
the Swiss Academy of Engineering Sciences (SATW), 
the Swiss National Energy Fund (NEFF) and 
the Swiss National Science Foundation (SNF) 
and is supported by the 
Assistant Secretary for Energy Efficiency
and Renewable Energy, 
Office of Building Technology Programs of the
U.S. Department of Energy, under Contract No.
DE-AC02-05CH11231.
I would like to thank these institutions for their generous support.

% =========================================================
\chapter{Legal}
\section{Copyright Notice}
GenOpt Copyright (c) \copyrightyears, The Regents of the University of California, through Lawrence Berkeley National Laboratory (subject to receipt of any required approvals from the U.S. Dept. of Energy).  All rights reserved.\\

If you have questions about your rights to use or distribute this software, please contact Berkeley Lab's Technology Transfer Department at \url{TTD@lbl.gov}.\\

{\bf NOTICE.} This software was developed under partial funding from the U.S. Department of Energy.  As such, the U.S. Government has been granted for itself and others acting on its behalf a paid-up, nonexclusive, irrevocable, worldwide license in the Software to reproduce, prepare derivative works, and perform publicly and display publicly.  Beginning five (5) years after the date permission to assert copyright is obtained from the U.S. Department of Energy, and subject to any subsequent five (5) year renewals, the U.S. Government is granted for itself and others acting on its behalf a paid-up, nonexclusive, irrevocable, worldwide license in the Software to reproduce, prepare derivative works, distribute copies to the public, perform publicly and display publicly, and to permit others to do so.

\section{License agreement}
GenOpt Copyright (c) \copyrightyears, The Regents of the University of California, through Lawrence Berkeley National Laboratory (subject to receipt of any required approvals from the U.S. Dept. of Energy).  All rights reserved.\\

Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
\begin{enumerate}
\item
Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
\item
Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
\item
Neither the name of the University of California, Lawrence Berkeley National Laboratory, U.S. Dept. of Energy nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
\end{enumerate}

THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT OWNER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.\\

You are under no obligation whatsoever to provide any bug fixes, patches, or upgrades to the features, functionality or performance of the source code ("Enhancements") to anyone; however, if you choose to make your Enhancements available either publicly, or directly to Lawrence Berkeley National Laboratory, without imposing a separate written license agreement for such Enhancements, then you hereby grant the following license: a  non-exclusive, royalty-free perpetual license to install, use, modify, prepare derivative works, incorporate into other computer software, distribute, and sublicense such enhancements or derivative works thereof, in binary and source code form.
